\chapter{Elaboration}

The problem of elaboration is to read an Alba source code file and build terms
and definitions in the intermediate language. In order to do this, overloaded
symbols have to be resolved and implicit arguments have to derived.

During the development of Alba I initially had completely underestimated the
complexity of the task. In other dependently type languages like
Idris~\cite{brady2013}, Agda~\cite{norell2007} and Lean~\cite{demoura2015} papers
have been written to describe the complex procedure.

In Alba we have the additional complexity that Alba shall be a language for
beginners and experts. The beginners should not be bothered with the complexity
of dependent types. Alba shall \emph{feel} like a simple functional programming
language. However the most powerful features of dependent types shall be
available in the language and the user should be able to grasp the power step by
step. I.e. the learning curve shall be flat and not steep.


The user terms are imcomplete. Not all binders are explicitly typed, implicit
arguments are usually ommitted, in local definitions types may be ommitted and
implicit arguments might be left out. It is the task of the elaborators to
derive complete terms from incomplete terms and issue understandable error
messages in case it cannot derive complete welltyped terms.

\paragraph{Metavariables}
Whenever the elaborator encounters an untyped binder or a missing implicit
argument, it introduces a metavariable. A metavariable is like a global function
which has a type but no definition. Possible arguments to the function are all
free variables in the context. I.e. a metavariable $\meta{m}$ has the type $\Pi
\vec{v}^\vec{V}. T_m$ where $\vec{v}$ are the free variables of the context.

Metavariables are instantiated by unification constraints. If the unifier
produces the constraint $\meta{m} \vec{v} \unify e$ or $\meta{m} \vec{v}
\unifysub e$ where all free variables in $\vec{v}$ are mutually distinct and $e$
contains only a subset of the free variables and $\Pi \vec{v}^\vec{V}. E \le \Pi
\vec{v}^\vec{V}.T_m$ where $E$ is the type of $e$, then the metavariable can be
instantiated by $\meta{m} := \lambda \vec{v}^\vec{V}. e$.

Unification constraints arise from comparing the type of an actual argument with
the expected formal argument type of the function term during a function
application. The type of the actual argument must be a subtype of the type of
the formal argument. The elaborator unifies the type of the actual argument with
the type of the formal argument. If any of the types contain metavariables, then
unification constraints can be produced.


Unification constraints of pattern type $\meta{m} \vec{v} \unify e$ where all
free variables are mutually distinct and the corresponding types can be unified
are immediately resolvable by $\meta{m} := \lambda \vec{v}. e$. But there are
other constraints which cannot be immediately resolved. These constraint must be
stored. The constraints can have dependencies i.e. they have a partial order.
Solution of the constraints must respect the order.

E.g. if the unification of the types of the left hand side and the right hand
side of $\meta{m} \vec{v} \unify e$ produces some remaining constraints, then
the resolution of this pattern type constraint has to be deferred until all
constraints on which it depends have been resolved.


\paragraph{Overloading} Global function or operator symbols and literal numbers
can be overloaded in Alba i.e. they can be ambiguous.

We allow ambiguity only if the number of explicit arguments is different or the
result type or some argument types of explicit arguments are different. Types
are different only if their head normal forms have a different global symbols in
the head position. Therefore the elaborator knows for each global symbol exactly
the positions of the explicit arguments which must be different and if the
result type must be different for overloaded symbols with the same number of
explicit arguments.

Whenever the elaborator encounter an overloaded symbol or literal, it knows the
number of explicit arguments and the result type. Partial application is
possible. Therefore not all explicit arguments might be present as actual
arguments.

If the elaborator cannot decide immediately by looking at the expected result
type which of the alternatives to use, it starts to introduce metavariables for
the expected types of the explicit arguments a starts the elaboration of the
arguments. If the elaboration of the arguments get stuck or does not provide
enough information, then the elaboration of the function application gets stuck.

If the elaboration of a function application gets stuck, then the elaborator
proceeds with the elaboration of the remaining subterm. The elaboration of the
remaining subterms might instantiate some metavariables such that the ambiguity
can be resolved.

But it might happen, that the ambiguity remains open. In that case we have to
try the possiblities.

Question: What happens if all possibilities are valid? Then the elaborator has
not enough information to resolve the ambiguity. Error message \emph{I cannot
decide which function to use.}

Question: What happens if all possibilities fail? They fail because some of the
constraints cannot be satisfied. Since the constraints are ordered we are only
interested in the possibilities which fail to satisfy the highest constraint in
the order. The source of unification problems is the compatibility of actual
arguments with formal arguments.






\section{Implicit Arguments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


In the type $\Pi x^A. B$ the variable $x$ can occur in $B$. If it does not occur
in the result type we simply write $A \to B$. The type $\Pi x^A. B$ is a
function type. We have the valid typing rule
$$
\rulev{
    \Gamma \vdash f: \Pi x^A. B

    \\

    \Gamma \vdash a: A
}
{
    \Gamma \vdash f a: B[x:= a]
}
$$

If $x$ occurs in $B$, then $a$ occurs in $B[x:=a]$. Therefore having the type of
$f a$ it is possible to \emph{infer} the argument $a$ by just looking at
$B[x:=a]$.


If the formal argument $x$ in the type $\Pi x^A. B$ occurs in the result type
$B$, then we call the argument $x$ \emph{potentially implicit}. In many cases a
potentially implicit argument is mandatory implicit as well.

The argument $x$ is \emph{mandatory implicit} in the type $\Pi x^A. B$, if it is
potentially implicit and one of the following conditions is satisfied.

\begin{enumerate}

\item
    $A$ is a kind i.e. either a syntactical kind or reduces to a syntactical
        kind. I.e. the normal form of $A$ has the structure $s$ or $\Pi
        \vec{z}^\vec{C}.  s$ where $s$ is a sort.


\item
    $B$ is a proposition i.e. $B: \Prop$. This implies that $\Pi x^A. B$ is a
        proposition as well.

\end{enumerate}


If a potentially implicit argument is not mandatory implicit, then the user has
the freedom to mark the argument as implicit or not. In all other cases the user
must mark the argument as implicit.

The following shows how explicit and implicit arguments appear in source code.


\begin{alba}
    all {x: A}: B               -- Implicit argument 'x'

    all (x: A): B               -- Explicit argument 'x'

    A -> B                      -- No implicit argument possible


    \ (x: A) ... := ...         -- Explicit argument 'x'

    \ {x: A} ... := ...         -- Implicit argument 'x'


    f a                         -- Explicit argument 'a'

    f {a}                       -- Implicit argument 'a'
\end{alba}
%
In an actual call the user has the freedom to provide an implicit argument or
ommit it. If the user provides the implicit argument, it has to be put into
braces.

A function type can have several implicit arguments in sequence. If the user
wants to supply a subsequence implicit argument, then the previous have to be
supplied as well.











\vskip 5mm
\section{Overloading}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Overloading means that the same name can be used for different
functions/constants. E.g. the operator $+$ can be addition for different type of
numbers like \code{Int} or \code{Float} or e.g. string concatenation. The
function \code{length} can be applied to a list, an array or a string argument.

Ambiguities shall be resolved in the context where the functions/constants are
used.

We have two different problems to solve:
\begin{enumerate}

\item Filter global ambiguous symbols before entering the context:
Functions/constants in the context with the same name shall be different either
for some explicit argument types or at the result type.

We shall not allow ambiguities in the global context which cannot be
distinguished by some explicit argument type or the result type.


\item Resolve ambiguities at usage: At the actual usage overloaded symbols shall
be resolvable.

\end{enumerate}

A type in normal form is either
\begin{enumerate}

\item a sort i.e. $\Prop$ or $\Any$.

\item a product $\Pi x^A. B$

\item a variable application $v a_0 a_1 \ldots$

\item a pattern match application
    $\case(\lambda \vec{y}^\vec{B} x^T. R, \vec{f}) \,\vec{a}\, e\, \ldots$
    %
    where $\vec{a}$ are the index arguments and $e$ is the object of the
    inductive type which is not a constructor expression.

\item a fixpoint application
    $\fix(x^T := t, k)\, \vec{a}\, e\, \ldots$
    %
    where $k = |\vec{a}|$ and $e$ is not a constructor expression.
\end{enumerate}
%
Due to the typing rules lambda abstractions cannot be types.





\subsection{Filter Overloaded Symbols}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsubsection{Comparing Types by Signatures}

We want to filter definitions before they enter the global context such that
only \emph{useful} overloadings are possible. The question: What is
\emph{useful}?

It is clearly useless to allow a symbol to be overloaded if both definitions
have the same type in normal form. This would create an ambiguity which can
never be resolved neither by type annotations, nor by actual argument types nor
by the required result type.

We allow overloadings only if the corresponding types differ significantly on
explicit arguments and the result type.

In order to compare types we convert them into a signatures. The atomic elements
of a signature are global names (e.g. \code{List}, \code{Int}, ...), sorts
represented by \code{Sort}, implicits
(represented by \code{I}) and unknowns (represented by \code{U}).

An element of a signature is either an atomic or an application.
\begin{alba}
    I
    U
    Int
    List U
    Array Nat
    Equal U U
    U U
\end{alba}

A signature is a list of elements or signatures where the last element is an
element. A signature has at least one entry representing the result type. The
optional entries before the last one represent the arguments. Examples
\begin{alba}
    [Int, Int, Int]
    [Float, Float, Float]
    [I, U, U, U]
    [I, U, List U, List U]
    [I, I, U, [U, U], U]
\end{alba}

When comparing signatures we do a structural comparison ignoring all implicit
argument positions. An unknown is not different from any other entry at the same
position. An unknown at the end is not different from any signature which has
more arguments.

Some examples of signatures
\begin{alba}
    (+): Int -> Int -> Int

        [Int, Int, Int]


    (+): String -> String -> String

        [String, String, String]


    identity: all {A: Any}: A -> A

        [I, U, U]
\end{alba}

Clearly the first two signatures are different. However the third is not
different from the first two. The significant part of the third signature is
\code{[U, U]} where the first unknown is neither different from \code{Int} nor
\code{String} and the second unknown is not different from the rest of both
other signatures because the second unknown is at the end.

The pipeline operator \code{|>} and the application operator \code{<|} have some
very general signatures.
\begin{alba}
    (|>): all {A: Any} {B: A -> Any} (a: A) (f: all x: B x): B a

        [I, I, U, [U, U], U]


    (<|): all {A: Any} {B: A -> Any} (f: all x: B x) (a: A): B a

        [I, I, [U, U], U, U]
\end{alba}

The type $\Pi A^\Any. A$ is the most general type and has the most general
signature.
\begin{alba}
    all {A: Any}: A

        [U]
\end{alba}



All other functions which act like attributes can be easily overloaded. Let's
look into the example of the \code{length} function.

\begin{alba}
    length: all {A: Any}: List A -> Int

        [I, List U, Int]


    length: all {A: Any}: Array A -> Int

        [I, Array U, Int]
\end{alba}

More example of signatures.
%
\begin{alba}

    Equal: all {A: Any}: A -> A -> Proposition

        [I, U, U, Sort]


    Plus {A: Any} (R: Endorelation A): Endorelation A

        [I, [U, U, Sort], U, U, Sort]


    naturalInduction
        {P: Predicate Natural}
        (zero: P zero) (step: all {i}: P i -> P (add1 i))
        {n}: P n

        [I, U, [U,U], I, U]


    listInduction
        {A: Any} {P: Predicate (List A)}
        : P [] -> (all {a l}: P l -> P (a :: l)) -> {l} -> P l

        [I, I, U, [U,U], I, U]


    lessEqualInduction {R: Endorelation Natural}
        :   (all {n}: R zero n)
            -> (all {n m}: LessEqual n m -> R n m -> R (add1 n) (add1 m))
            -> all {n m}: LessEqual n m -> R n m

        [I, U, [I, I, LessEqual U U, U, U], LessEqual U U, U]
\end{alba}

Note that everything which is neither an implicit argument nor a global function
nor a sort is \emph{unknown}. Applications of implicit arguments to any
arguments are \emph{unkown} in the signature.





\subsubsection{Computed Types}

In Alba types can be computed as well. Two examples to demonstrate a computed
type.
%
\begin{alba}
    (inspect n case
        zero := True
        add1 i := False)

    (inspect n case
        zero := Int
        add1 i := List Int)
\end{alba}
%

Since computed types can have any forms we consider them as unknown in
signatures.










\subsection{Resolve Ambiguities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

During the elaboration of a term, the possible ambiguities have to be resolved.
There are only two sources for ambiguity:
%
\begin{enumerate}

\item Overloaded global symbols like $+$ and \code{length}

\item Literal numbers (they can have type \code{Int}, \code{Float},
\code{Natural}, etc.).

\end{enumerate}








\section{Metavariables and Unification}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsection{Metavariables}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We use metavariables or holes to describe terms which have to be derived by
unification. There are some reasons to introduce metavariables.

\begin{itemize}

\item Implicit arguments

\item Untyped variables

\item Required types for terms which do not have an explicit required type.

\item Convert the type of an untyped variable to a function type.

\end{itemize}


The context of elaboration has 3 types of variables.

\begin{itemize}

\item Constants: All variables which have already been in the context at the
start of elaboration.

\item Free variables: All variables introduced to start the elaboration of an
abstraction, a function type or a where expression.

\item Bound variables: Variables shifted into the context during unification to
unify abstractions, function types or where expressions.

\end{itemize}

In the elaboration procedure of the lean theorem prover described in
\cite{demoura2015} metavariables are defined in a way such that they can only
instantiated by closed terms i.e. by terms which do not contain neither free
variables nor bound variables.

This can be achieved by using metavariables of the form $\meta{M}: \Pi
\vec{x}^{\vec{A}}.T$ where $\vec{x}$ is the collection of all free variables and
using the metavariable only in the form $\meta{M} \vec{x}$.

Whenever we try to unify
$$
    \meta{M} \vec{x} \le T
$$
we get the solution
$$
\begin{array}{lll}
    \meta{M} &:=& \lambda \vec{x}^{\vec{A}}. T

    \\

    \meta{M} \vec{x} &\leadsto& T
\end{array}
$$

It is not yet clear if such a form of introducing metavariables is really
necessary. But this form has the advantage to include in each introduction of a
metavariable the variables on which it can depend. The unification problem
$\meta{M} \vec{x} \le T$ indicates that $T$ is allowed to depend only on the
variables $\vec{x}$. If $T$ contains more variables, we can reject the
instantication of $\meta{M} := \lambda \vec{x}^{\vec{A}}. T$ because the
definition term is not closed.

If we don't use this scheme to introduce metavariables we have to make sure that
metavariables are not instantiated inconsistently.




\subsection{Unification}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Unification is usually triggered during elaboration of an application. The type
of the actual argument must be a subtype of the argument required from the
function term. Therefore a unification problem usually starts with
$$
    A \le R
$$
where $A$ is the type of the actual argument and $R$ is the required type of the
argument.















\section{Elaboration of Terms}
%====================================





\subsection{Function Call with Unknown Function Type}
%----------------------------------------------------


\begin{alba}
    f a

    -- where
    f: ?F v     -- 'v' is array of free variables
\end{alba}

In the best case the type of the argument is known $a: A$ and the result type is
known $f a: R$. In this case we can use $A \to R$ as the type of $f$.

We can try to use a dependent product for the type of $f$. We could try
$$
    f: \Pi x^{\meta{A} \vec{v}}. \meta{B} \vec{v}x
$$
%
This results in the unification constraints
$$
\begin{array}{lll}
    \meta{A} \vec{v}
    &\unify&
    A

    \\

    \meta{B} \vec{v} a
    &\unify&
    R
\end{array}
$$
%
The first one can be immediately resolved by $\meta{A} := \lambda \vec{v}. A$.
The second one is not of the pattern type, because $\vec{v} a$ are not mutually
distinct free variables. Therefore it cannot be resolved immediately. It might
be resolvable in the course of the elaboration. However at that point the
application $f a$ cannot be safely made because it is only typesafe under the
unresolved unification constraint.




\vskip 5mm
\subsection{Function Call}
%-------------------------


\begin{alba}
    f a b c ...
\end{alba}
where the function term \code{f} is one of the following

\begin{itemize}
\item
    Global or local name
\item
    Pattern match
\item
    Function abstraction
\end{itemize}

The function term can be ambiguous e.g. if it is an overloaded global name.

In case of ambiguity, it can be resolved by looking at the call context (number
of explicit arguments), the signature of the expected type and the argument
types. The overloaded symbol has a set of signatures. The signatures tell the
criteria on which the decision can be made.

First we filter out the signatures which fit the call context.



\subsection{Where Expression}
%-----------------------------

\begin{alba}
    body where
        ...
        h ... := ...
        g ... := ...
        f ... := ...
\end{alba}

\code{f} is the outermost local definition. It has to be introduced into the
context as a free variable.

If the local definition is completely untyped e.g.
\code{f x y := x + y} we cannot infer a type of \code{f} and we have to
introduce a metavariable for its type.

The signature might be given partially of fully. We analyze the signature of $f$
from left to right until no more type annotation is given. If the signature is
given only partially, we introduce a metavariable for the remaining untyped
arguments and the result type.

We elaborate the body only if the signature of a local definition is given
fully.

We most usual case is that the signature is not given or given only partially
and the type shall be inferred from its use in the inner local definitions or in
the body.





\subsection{Pattern Match}
%-------------------------


\begin{alba}
    inspect e case
        {\ {x} z :=
            {- optional elimination function -}
        }
        c0 y00 y01 ... := d0
        c1 y10 y11 ... := d1
        ...
\end{alba}





\begin{enumerate}
\item
    Collect constructors: Each case begins with a constructor. If we collect
    all constructors we can find the candidate types for the inductive type.

    Possible errors are:

    \begin{itemize}
        \item Some of the \code{ci}s are not constructors.
        \item The set of constructors do not belong to the same inductive type.
    \end{itemize}

    If we get one (or more) inductive type, we get it in the form
    $$
        \Inductive_i (\ind)
        :
        \Pi \vec{p}^\vec{P} \vec{x}^\vec{A} . s
    $$
    i.e. we do not yet know the parameters and the indices.

\item
    Build the inspected expression \code{e} with an (or more) expected inductive
    type. We need metavariables for the parameters and the indices of the
    inductive type, because the set of constructors identify only the
    inductive type, but not the parameters and the indices.

    I.e. we get an expression of the form
    $$
        e: \Inductive_i (\ind) \vec{q} \vec{a}
    $$
    where $\vec{q}$ are the actual parameters and $\vec{a}$ are the actual
    indices of the inductive type.

\item
    Infer or build the elimination function. The elimination function must have
    the type
    $$
        \Pi \vec{x}^\vec{A} z^{\Inductive_i (\ind) \vec{q} \vec{x}} . s_R
    $$
    with the actual parameters $\vec{q}$ which we have derived in the previous
    step. The sort $s_R$ of the result type is still unknown.

    If the elimination function is explicitly given, we can assign (or check)
    the types of $\vec{x}$ and $z$ and elaborate it.

    If the elimination function is not available we introduce a metavariable
    $\meta{R}$ of the correct type of the elimination function and unify
    $\meta{R}
    \vec{q} \vec{a}$ with the expected result type of the whole inspect
    expression.
\end{enumerate}








\section{Elaboration Primitives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Each term which we are going to build has a requirement. The requirement can be
either empty, or the term must represent a type or the type of the term must be
a subtype of a required type.

We can build terms without any required type. Such terms do not satisfy any
requirement. If we unify a the type of a term with a requirement, then the
term satisfies the requirement and that satisfaction is stored within the
elaborated term.

We can also unify an elaborated term directly with some requirement. Then the
elaborated term is a subtype of the requirement. This subtype satisfaction is
stored as well in the elaborated term.


\begin{description}

\item [Metavariable] We need a primitive to create a metavariable. In the
    metavariable we store the reason why it has been created. The metavariable
        serves either as a type for an untyped variable, as the type of a
        toplevel expression to build or as a placeholder for an implicit
        argument of a function call.

        In the last case we have a type for the metavariable because every
        implicit argument has a type.


\item [Sort]
    Build the sort and unify its type with the required type. If the unification
        is successful, mark the term to satisfy its required type.

        Note: A sort cannot be applied to arguments. If there are explicit
        arguments, report an error.

\item [Variable]
    A variable can be applied to arguments. Therefore the required type is valid
        only as a result type for the whole application, not for the variable
        alone.

        First we build the variable without any requirement. With the build we
        get the signature of its type. The signature of its type shows on how
        many arguments the variable can be applied, the positions of the
        implicit arguments and the result type.

        From the signature we can skip the actual arguments and all implicits
        which occur before. This gives us a result signature in this specific
        call context. Note: If the result type in the signature is unknown it is
        valid that there are more explicit arguments than the signature shows.

        The actual result signature in this specific call context can be
        compared with the signature of the required type. From the comparison we
        see, if the actual variable can return an object of the corresponding
        result type and if more implicit arguments have to be entered after the
        last explicit argument.

        If there are no explicit actual arguments, we generate the possibly
        remaining implicit arguments and unify the actual result type with the
        required result type.


\item[Application] First we elaborate the function term with the requirement for
    the whole application and the number of explicit arguments incremented by
        one.

        We look into the signature of its type. It has to accept at least one
        explicit arguments.

        If this is possible we generate metavariables for the implicit arguments
        before the next explicit argument and apply the function term to the
        implicit arguments.

        The we extract the expected type of the next explicit argument and
        generate a requirement for the actual argument.

        We build the argument term with the requirement. The we apply the
        function term to the argument.

        Then we have check, if more implicit arguments are needed to satisfy the
        requirement of the application.

\item [Typed Term] First we elaborate the type of the typed term with the
    requirement of being a type. Then we unify the elaborated type with the
    requirement.

    Then we elaborate the term of the typed term with the type of the typed
    term as a requirement.

    After doing this successfully we can form the typed term which is guaranteed
    to satisfy the requirement of the whole term.

    If the typed term is applied to actual arguments, then the unificatioin of
    the elaborated type with the requirement cannot be done. If just can be
    checked, if the elaborated type is consistent with the requirement after
    application of some more arguments.


\item [Product] A product cannot be applied to arguments. If there are actual
    arguments, then an error has to be reported.

    The requirement has to be a sort. If not, an error has to be reported.

    From the requirement of the whole product we can generate requirements for
    the argument type and the result type.

    If the requirement for the product is a proposition, then the argument type
    can be an arbitrary type and the result type has to be a proposition as
    well.

    If the requirement for the product is an impredicative universe, then the
    requirement for both the argument type and the result type is the same
    impredicative universe.

    If the requirement for the product is an arbitrary type, then the argument
    type and the result type can be arbitrary types.

    We elaborate the argument type. Then we add a new free variable with the
    argument type. Next we elaborate the result type.

    Finally we generate the product type by abstracting out the ultimate free
    variable.


\item [Abstraction] An abstraction can be applied to arguments. If there are
    arguments, we try to elaborate a redex.

    \begin{comment}
        blalbkakljdalk dlskj
    \end{comment}



\item [Where Expression]

\item [Pattern Match]

\item [Fixpoint]

\end{description}











\section{Where Expressions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsubsection{Example}

\begin{alba}
    1 + f 10
    where
        f x := 2 * x + 25
\end{alba}

We need a metavariable for the type of $f$ and introduce $f$ as a free variable
into the context, i.e. we need the context
$$
    \Gamma, {\meta{T}_f}^{\Any_1}, f^{\meta{T}_f}
$$

In this context start to analyze \code{1 + f 10}. $+$ is an overloaded symbol
for the various numbers and other overloadings. After seeing the number $1$ as
the first argument, only addition of numbers remains. But it still remains
ambiguous.

In trying to elaborate the second argument \code{f 10} we have as the required
type one of the possible numbers. Now the elaboration becomes stuck, because the
elaboration of \code{10} is ambiguous and we don't want to nest ambiguities.

So we have to elaborate the local definition next. We have to add a metavariable
$\meta{T}_x$ for the type of $x$ and the free variable $x$ to the context.
$$
    \Gamma, {\meta{T}_f}^{\Any_1}, f^{\meta{T}_f}, {\meta{T}_x}^{\Any_1}, x^{\meta{T}_x}
$$

The we elaborate \code{2*x + 25} with any of the possible number types as the
required type. This elaboration is unique for any of the number types and
assigns the corresponding number type to the metavariable $\meta{T}_x$.

Elaborating the abstraction we get
$$
    (\lambda x^N. 2x + 25): \Pi x^N. N
$$
where $N$ is the corresponding number type. Now we get the assignment $\meta{T}_f
:= \Pi x^N. N$ and can continue with the elaboration of \code{f 10}.


\vskip 5mm
\subsubsection{General Case}

Draft: $\equiv, \approx, \cong, \simeq, \doteq, \precapprox, \precsim, \lesssim,
\gtrsim$.



\begin{alba}
    body
    where
        a := def   -- 'def' can be '\ x y ... := e'

    body
    where
        f x y .. := def
\end{alba}
