\section{Relation to Other Typed Lambda Calculi}
\label{sec:lang-other-typed}


In the section \emph{Intermediate
  Language}~\ref{sec:lang-intermediate-language} a typed lambda calculus has
been presented which serves as an intermediate language for Alba programs. The
presented lambda calculus is fairly general. In this chapter we explain, how
other typed lambda calculi are contained as special cases.

In particular we demonstrate that all typed lambda calculi of the Barendregt
cube are contained as special cases.


\subsection{Barendregt Cube}

For all typed calculi in the Barendregt cube we have no inductive types, no
abstract types, no pattern match and no fixpoint constructs. Furthermore we
have only the sorts $\Prop$ and $\Any$. I.e. our terms
are defined by the grammar
$$
\begin{array}{lllll}
  t &:= & \Prop \mid \Any  & \text{sorts}
  \\
    & \mid & x & \text{variable}
  \\
    & \mid & f \, a & \text{application}
  \\
    & \mid & \lambda x^A. e & \text{abstraction}
  \\
    & \mid & \Pi x^A.B & \text{dependent product}
\end{array}
$$.

We have no subtyping relation and only one axiom
$$
 [] \vdash \Prop : \Any
$$

Although it sounds a little bit funny any type $T$ with $ T : \Prop$ is a
\emph{normal} type i.e. \emph{normal} types are propositions.

The power and expressibility of a typed lambda calculus is determined mainly
by the allowed rules to form function types i.e. dependent products. We need
the generic rule
$$
\rulev
{
  \Gamma \vdash A : s_1
  \\
  \Gamma, x^A \vdash B : s_2
}
{ \Gamma \vdash \Pi x^A . B : s_2}
$$

\begin{tabular}{| l | l | l | l |}
  \hline
  & $s_1$ & $s_2$ &
  \\
  \hline
  1. fun       & $\Prop$ & $\Prop$ & functions from objects to objects
  \\
  2. polyfun   & $\Any$  & $\Prop$ & polymorphic functions from types to objects
  \\
  3. polytype  & $\Any$  & $\Any$  & polymorphic functions from types to types
  \\
  4. deptype   & $\Prop$ & $\Any$  & functions from objects to types
  \\
  \hline
\end{tabular}

In the typed calculus of the intermediate language the typing rule for
products is a littly bit more restrictive, but by combining the product rule
and the subtype rule with the sorts $\Prop$, $\Any$ and $\Box$ the same effect
as the above product rules can be achieved.

Depending on which product rules we allow we get the main typed calculi of the
Barendregt cube by
\begin{enumerate}
\item Simply typed lambda calculus (also called $\lambda_\to$): rules =
  \{fun\}

\item System $F$ (also called $\lambda 2$): rules = \{fun, polyfun\}

\item System $F_\omega$ (also called $\lambda 2\omega$): rules = \{fun,
  polyfun, polytype\}

\item Calculus of Constructions (also called $\lambda C$): rules = \{fun,
  polyfun, polytype, deptype\}
\end{enumerate}

These are the most interesting calculi of the Barendregt cube. In total there
are $2^3 = 8$ calculi for all other possible combinations of the rules which
contain at least the rule fun.





\subsection{Church Encoding of Inductive Types}

Since we don't have inductive types, we need some other mechanism to encode
nontrivial functions. In order to represent interesting features we need at
least system $F$ or sometimes system $F_\omega$. The expressive power of
simply typed lambda calculus is rather limited.

In order to find Church encodings of inductive types we use the following
guidelines
\begin{enumerate}
\item An object of an inductive type is encoded as a function with one
  argument for each constructor plus appropriate type arguments. One type
  argument is always the result type of the operation.

\item The signature of the argument representing a constructor can be found by
  looking at the signature of the constructor.

\item Recursive argument types in constructors are replaced by arguments of
  the result type of the operation.

\item The type of an object representing a Church encoding of an object of an
  inductive type looks like the type of a folding function.
\end{enumerate}


\subsection{Booleans}

With inductive types we define booleans as
\begin{alba}
  class Boolean: Any create
    true:  Boolean
    false: Boolean
\end{alba}
%
with the folding function
%
\begin{alba}
  fold (b:Boolan) (A:Any) (x y:A): A :=
    inspect b case
      true :=
         x
      false :=
         y
\end{alba}


Since \code{Boolean} has two constructors we encode boolean values as
functions with three arguments. The first argument is the result type and the
other two arguments represent the two constructors. Since the constructors
have no arguments, the arity of the corresponding arguments is 0.

$$
\begin{array}{lllll}
  B & :=
  & \Pi A^\Prop . A \to A \to A
  &:
  & \Prop
  \\
  \text{true} &:=
  & \lambda A x y . x
  &: & B
  \\
  \text{false} &:=
  & \lambda A x y . y
  &: & B
\end{array}
$$

The type of a boolean value is a function which receives a type, two objects
of that type and returns an object of that type.

The function $\text{true}$ returns the first of the two objects, the function
$\text{false}$ returns the second of the two objects. A boolean value encodes
a decision between two alternatives.  $\text{true}$ decides for the first
alternative and $\text{false}$ decides for the second.



\subsection{Pairs}

With inductive types we define pairs as
\begin{alba}
  class (,) (A B: Any): Any create
    (,): A -> B -> (A,B)
\end{alba}
%
with the folding function
%
\begin{alba}
  fold A B (p: (A,B)) (C:Any) (f: A -> B -> C): C :=
    inspect p case
      (a,b) :=
         f a b
\end{alba}

Using the folding function we can encode the projections on the first and the
second element of the pair.
\begin{alba}
  first A B (p: (A,B)): A :=
     p.fold (\ x y := x)

  second A B (p: (A,B)): B :=
     p.fold (\ x y := y)
\end{alba}

From that we derive the Church encoding

$$
\begin{array}{lllll}
  P & :=
  & \lambda A B . \Pi C^\Prop . (A \to B \to C) \to C
  &:
  & \Prop \to \Prop \to \Prop
  \\
  (,) &:=
  & \lambda A B x y C f . f x y
  &: & \Pi A^\Prop B^\Prop x^A y^B.  P \, A \, B
  \\
  \text{first} &:=
  & \lambda A B p . p A (\lambda x y. x)
  &: & \Pi A^\Prop B^\Prop. P\, A\, B \to A
  \\
  \text{second} &:=
  & \lambda A B p . p A (\lambda x y. y)
  &: & \Pi A^\Prop B^\Prop. P\, A\, B \to B
\end{array}
$$






\subsection{Natural Numbers}


The following defines the inductive type of natural numbers in Alba.
\begin{alba}
  class Natural: Any create
    0: Natural
    succ: Natural -> Natural
\end{alba}

The folding function we call \code{iterations} because it does a number of
iterations.

\begin{alba}
  iterations (n:Natural) (A:Any) (f:A -> A) (s:A): A :=
      -- iterations is in that case a better name for the folding function
    inspect n case
      0 :=
        s
      succ k :=
        f (fold k f s)
\end{alba}

Using the folding function it is easy to define the standard arithmetic
operations.


\begin{alba}
  (+) (a b: Natural): Natural :=
    a.iterations succ b

  (*) (a b: Natural): Natural :=
    a.iterations (\ s := s + b) 0

  (^) (a b: Natural): Natural :=
    b.iterations (\p := p * a) (succ 0)
\end{alba}

\noindent The Church encodings:
$$
%
\begin{array}{lllll}

  \Natural & :=
  & \Pi A^\Prop . (A \to A)  \to A \to A
  &:
  & \Prop

  \\

  0 & :=
  & \lambda A f s . s
  &:
  & \Natural

  \\

  \text{succ} & :=
  & \lambda k A f s . f (k f s)
  &:
  & \Natural \to \Natural

  \\

  (+) & :=
  & \lambda a b. a\, \Natural\,\text{succ}\, b
  &:
  & \Natural \to \Natural \to \Natural

  \\

  (*) & :=
  & \lambda a b. a\, \Natural\, (\lambda s. s + b) 0
  &:
  & \Natural \to \Natural \to \Natural

  \\

  (\caret) & :=
  & \lambda a b. b\, \Natural\, (\lambda p. p * a) (\text{succ}\, 0)
  &:
  & \Natural \to \Natural \to \Natural
\end{array}
$$


We can use the folding function to decide if a natural number is zero.

\begin{alba}
  is_zero (n: Natural): Boolean :=
    n.iterations (\_ := false) true
\end{alba}

A simple check shows that this function is correct. However it is rather
inefficient, because it iterates over the complete structure, throwing away
all previous results.

With Church encoding the function looks like
$$
0? := \lambda n. n B (\lambda x . \text{false}) \text{true} : \Natural \to B
$$


In a similar manner we can define the predecessor function based on the
folding function.

\begin{alba}
  predecessor (n: Natural): Natural :=
    first (n.iterations (\p := (p.second, 1 + p.second)) (0, 0))
\end{alba}

We iterate over the natural number by generating a pair. The first component
of the pair is the predecessor and the second component is the number. In each
iteration step the first component is thrown away, only the intermediate
number is used to form a new predecessor-number pair. At the end of the
iteration we throw away the number part of the pair.

The same function in Church encoding
$$
\text{pred} :=
\text{first}\,
((\lambda n . n \Natural \lambda p. (\text{second}\, p, \text{succ}\,
(\text{second}\, p))
(0,0))
: \Natural \to \Natural
$$



\subsection{Ackermann Function}

It is quite instructive to see that the Ackermann function can be defined with
Church encodings.

The Ackermann function is defined by the recursive equations

$$
A \, m \, n :=
\begin{cases}
  A\, 0\, n &:= 1 + n
  \\
  A\, (1+k)\, 0 &:= A\, k\, 1
  \\
  A\, (1+k)\, (1+o) & := A\, k\, (A\, (1+k)\, o)
\end{cases}
$$
%
Wilhelm Ackermann defined the function slightly different, but the above
definition is equivalent and today used in the literature as the definition of
\emph{the} Ackermann function.

It is known that this function is not primitive recursive because it grows
faster than any primitive recursive function. Wilhelm Ackermann invented this
function to prove that there are computable functions which are not primitive
recursive.

We define the Ackermann function in Alba by using the folding function

\begin{alba}
  g (f:Natural -> Natural) (o:Natural): Natural :=
    (1 + o).iterations f 1

  ack (m:Natural): Natural -> Natural :=
    m.iterations g successor
\end{alba}

or in Church encoding
$$
\begin{array}{lllll}
  g &:=
  & \lambda f o . (\text{succ}\, o)\, \Natural f (\text{succ}\, 0)
  &:& (\Natural \to \Natural) \to \Natural \to \Natural

  \\

  \text{ack} &:=
  & \lambda m. m (\Natural\to\Natural)\, g\, \text{succ}
  &:& \Natural \to \Natural \to \Natural

\end{array}
$$

We prove that the implementation satisfies the recursion equations. We first
start with the observation

\begin{alba}
  ack (1 + k) o  = (1 + k).iterations g succ o
                 = g (k.iterations g succ) o
                 = g (ack k) o
\end{alba}

The we prove the 3 equations which have to be satisfied by the Ackermann
function
\begin{enumerate}
\item
  \begin{alba}
    ack 0 n  = 0.iterations g succ n
             = succ n
  \end{alba}


\item
  \begin{alba}
    ack (1 + k) 0  = g (ack k) 0               -- see above
                   = 1.iterations (ack k) 1    -- def g
                   = ack k 1
  \end{alba}


\item
  \begin{alba}
    ack (1 + k) (1 + o) = g (ack k) (1 + o)             -- see above
                        = (2 + o).iterations (ack k) 1  -- def g
                        = ack k ((1 + o).iterations (ack k) 1)
                        = ack k (g (ack k o))           -- def g
                        = ack k (ack (1 + k) o)         -- see above
  \end{alba}

\end{enumerate}

It seems that the category of primitive recursive functions is not an
important function category in typed lambda calculi. The simply typed lambda
calculus is not powerful enough to represent all primitive recursive
functions. The next more powerful calculus system F is already powerful enough
to represent more than primitiv recursive functions.

Furthermore it is interesting that adding polymorphic functions to the simply
typed calculus add enormous expressive power to the calculus. Higher order
functions (which is standard in any lambda calculus) and polymorphic functions
are sufficient to represent functions like the Ackermann function.




\subsection{Lists}

The inductive type of lists and the corresponding folding function is defined
in Alba as

\begin{alba}
  class List (A: Any): Any create
    []: List A
    (^): A -> List A -> List A

  fold A (a: List A) (B:Any) (f:A -> B -> B) (s:B): B :=
    inspect a case
      [] :=
        s
      h ^ t :=
        f h (fold t f s)
\end{alba}

A corresponding Church encoding

$$
\begin{array}{lllll}
  L &:=
  & \lambda A . \Pi B^\Prop . (A\to B\to B) \to B \to B
  &:& \Prop \to \Prop

  \\

  [] &:=
  & \lambda A B f s . s
  &:& \Pi A^\Prop . L A

  \\

  (\caret) &:=
  & \lambda A h t B f s . f h (t B f s)
  &:& \Pi A^\Prop . A \to L A \to L A
\end{array}
$$


\subsection{Binary Trees}

Binary trees can be encoded in the same manner.

\begin{alba}
  class Tree (A:Any): Any create
    empty: Tree A
    node : Tree A -> A -> Tree A -> Tree A

  fold A (t:Tree A) (B:Any) (f:B -> A -> B -> B) (s:B): B :=
    inspect t case
      empty :=
        s
      node l a r :=
        f (fold l f s) a (fold r f s)
\end{alba}

$$
%
\begin{array}{lllll}
  T &:=
  & \lambda A. \Pi B^\Prop. (B\to A\to B\to B) \to B \to B
  &:& \Prop\to\Prop

  \\

  \text{empty} &:=
  & \lambda A B f s . s
  &:& \Pi A^\Prop . T A

  \\

  \text{node} &:=
  & \lambda A l a r B f s . f (l B f s) a (r B f s)
  &:& \Pi A^\Prop . T A \to A \to T A \to T A
\end{array}
$$


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main_alba_design"
%%% End:
