\section{Arbitrary Precision Arithmetic}

We have a base $B$ and digits $0 \le d < B$ i.e. the greatest digit is $B-1$.
\begin{itemize}

\item The digit sequence $(d_0 d_1 d_2 \ldots d_{n-1})_B$ is interpreted as the number
  $(\ldots((d_0 B + d_1)B + d_2) + \ldots) + d_{n-1}$ i.e.

    $$ d_0 B^{n-1} + d_1 B^{n-2} + \ldots + d_{n-2} B + d_{n-1} $$


\item The result of adding two digits is at most $2B-2$.

\item The result of adding two digits and a carry ($0$ or $1$) is at most
  $2B-1$.

\item For a two digit sequence where the first digit is $1$ we get
  $(1 d)_B = 1 B + d \le 2 B - 1$

\item For the four digits $a,b,c,d$ we get $a b + c + d < B^2$ because of $a b
  \le (B - 1)^2 = B^2 - 1 - (2B-2)$ and $c + d \le 2B-2$. Furthermore $(a b +
  c + d)/B < B$ is valid.
 \end{itemize}


Primitive operations
\begin{enumerate}
\item Addition and substraction of one-place integers, giving a one-place
  answer and a carry.

\item Multiplication of a one-place integer with another giving a two-place
  answer.

\item Division of a two-place integer by a one-place integer, provided that
  the quotient is a one-place integer, and yielding also a one-place
  remainder.

\end{enumerate}






\subsection{Division}



\subsubsection{Basics}

In order to define the division algorithm we have to define integer division and
prove some properties of integer division.

\begin{definition}

    We write the integer division of the nonegative number $a$
    divided by the positive number {b} as

    $$\intdiv{a}{b}$$

    \noindent $\intdiv{a}{b}$ is the number $n$ satisfying the specification

    $$
    nb \le a \le (n+1) b - 1
    $$

    Note that the specification defines the number $n$ uniquely.
\end{definition}

\begin{theorem}\label{intdiv-theorem}
    Any number $n$ satisfying the specification
    %
    $$n b \le a \le (n+1) b - 1$$
    %
    for $0 \le a$ and $0 < b$ satisfies the condition
    %
    $$ n b f \le a f + i \le (n+1) b f - 1 $$
    %
    for all $0 \le i < f$. I.e.
    %
    $$
    \intdiv{a}{b} = \intdiv{af}{bf} = \intdiv{af + i}{bf}
    $$
    is valid.

    \begin{proof}
        We prove this claim by the following chain of order preserving
        transformations.
        %
        $$
        \begin{array}{lllll}
            n b &\le& a &\le& (n+1) b - 1
            \\
            nbf &\le& af &\le& (n+1) b f - f
            \\
            nbf + i &\le& af + i &\le& (n+1) b f - (f - i)
            \\
            nbf     &\le& af + i &\le& (n+1) b f - 1
        \end{array}
        $$
        %
        The last step is valid because of $0 \le i < f$. The first condition is
        equivalent to
        $$ n = \intdiv{a}{b}$$
        The last condition is equivalent to
        $$ n = \intdiv{af + i}{bf}.$$
    \end{proof}
\end{theorem}


\begin{theorem}\label{intdiv-theorem2}
    For all $0 \le a$ and $0 < b$ we have
    $$
        \frac{a}{b} - 1 < \intdiv{a}{b}
    $$

    \begin{proof}
        Let $n := \floor{a/b}$. This implies that $n$ satisfies
        $$
            nb \le a \le (n + 1) b - 1
        $$
        %
        We use the right inequality and derive
        $$
        \begin{array}{lll}
            a &\le& (n + 1) b - 1
            \\
            a/b  &\le&   n + 1 - 1/b
            \\
            a/b  &<&   n + 1
            \\
            a/b - 1  &<&   n
            \\
            a/b - 1 &<&   \intdiv{a}{b}
        \end{array}
        $$
    \end{proof}
\end{theorem}


\subsubsection{Basic Division}

In this subsection let $u = (u_0 u_1 \ldots u_n)_B$ and $v = (v_1 v_2 \ldots
v_n)_B$ and $u / v < B$.
%
We want to find $q := \floor{u / v}$.

We start with the initial guess
$$
\hat{q} := \text{min}( \intdiv{u_0 B + u_1}{v_1}, B - 1)
$$


\begin{theorem}[Initial guess is upper bound]
    \label{initial-guess-upper-bound}
    The initial guess is an upper bound for the correct value i.e. $q \le
    \hat{q}$.

    \begin{proof}
        Since $q < B$ because of $u/v < B$ the claim is evident for $\hat{q} = B
        - 1$. Therefore we treat the case $\hat{q} < B - 1$.

        We always have
        $$
            (u_2 \ldots u_n)_B < B^{n-1}
        $$
        by definition of positional numbers. If we use the theorem about integer
        divisions~\ref{intdiv-theorem} with $f := B^{n-1}$ and $i := (u_2 \ldots
        u_n)_B$ we get
        $$
        \begin{array}{lll}
            \hat{q}
            &:=&
            \intdiv{u_0 B + u_1}{v_1}
            \\
            \\
            &=&
            \intdiv{(u_0 B + u_1) B^{n-1}}{v_1 B^{n-1}}
            \\
            \\
            &=&
            \intdiv{(u_0 B + u_1) B^{n-1} + (u_2 \ldots u_n)_B}{v_1 B^{n-1}}
            \\
            \\
            &=&
            \intdiv{u}{v_1 B^{n-1}}
        \end{array}
        $$

        In general we have $\floor {u / w_1} \le \floor{u / w_2}$ if $w_2 \le
        w_1$ because dividing by a bigger value leads to a smaller result. Since
        $v_1 B^{n-1} \le v$ we get
        $$
            \intdiv{u}{v} \le \intdiv{u}{v_1 B^{n-1}}
        $$
        which proves the claim $q \le \hat{q}$.
    \end{proof}
\end{theorem}


\begin{theorem}[Initial guess maximal error is 2]
    The initial guess $\hat q$ and the correct value $q$ are at most 2 apart
    i.e. $\hat q - q \le 2$ as long as $\floor {B/2} \le v_1$ is valid.

    \begin{proof}

        Instead of proving $\floor{B/2} \le v_1 \imp \hat q - q \le 2$
        we prove the contrapositive $3 \le \hat q - q \imp v_1 < \floor{B/2}$.
        So let's assume $q + 3 \le \hat q$ and prove the following subclaims.

        \begin{enumerate}
            \item $v_1 \le v / B^{n-1}$:

                By definition of positional numbers we have $v_1 B^{n-1} + (v_2
                \ldots v_n)_B = v$ which implies $v_1 B^{n-1} \le v$. Dividing
                both sides by $B^{n-1}$ proves the subclaim.

            \item $B^{n-1} < v$:

                From the previous subclaim we conclude $v_1 B^{n-1} \le v$ and
                since $v_1$ is different from zero we get $B^{n-1} \le v$.

                Now assume by way of contradiction $B^{n-1} = v$. In that case
                $\floor{u / v}$ and $\floor{u_0 B + u_1 / v_1}$ are the same and
                therefore $\hat q$ and $q$ must be the same (note that
                $\floor{u/v} < B$). This contradicts our assumption that $\hat
                q$ and $q$ are $3$ apart.

            \item $u/v - 1 < q$:

                This is an immediate consequence of the definition of $q$ and
                the theorem~\ref{intdiv-theorem2}.

            \item $\hat q \le u /(v - B^{n-1})$:

                Since $\hat q := \text{min}(\floor{u_0 B + u_1 / v_1})$ and
                $\floor{u_0 B + u_1 / v_1} = \floor {u / v_1 B^{n-1}}$ (see
                reasoning in theorem~\ref{initial-guess-upper-bound}) we get
                $\hat q \le \floor{u /(v_1 B^{n-1}})$ which implies
                $\hat q \le u / (v_1 B^{n-1}) $. Together with $v < v_1 B^{n-1}
                + B^{n-1}$ we get the subclaim.

            \item $2 (v_1 - 1) \le B - 4$:

                We prove this by the following chain:
                $$
                \begin{array}{llll}
                    q + 3 &\le& \hat q      &\text{assumption}

                    \\ \\

                    u/v - 1 + 3 &<& u / (v - B^{n-1})
                    &\text{subclaim 3 and 4}

                    \\
                    \\

                    2 &<& \frac{u}{v - B^{n-1}} - u/v

                    \\
                    \\

                    2 &<& \frac{u v - u (v - B^{n-1})}{v (v - B^{n-1})}

                    \\
                    \\

                    2 &<& \frac{u}{v} \frac{v}{v - B^{n-1}}

                    \\ \\

                    2 (v/B^{n-1} - 1) &<& u/v

                    \\ \\

                    2 (v_1 - 1) &<& u/v  &\text{subclaim 1}

                    \\ \\

                    2 (v_1 - 1) &\le& \intdiv{u}{v}
                    &\text{left hand side is an integer}

                    \\ \\

                    2 (v_1 - 1) &\le& q &\text{definition of $q$}

                    \\ \\

                    2 (v_1 - 1) &\le& \hat q - 3 &\text{assumption}

                    \\ \\

                    2 (v_1 - 1) &\le&  B - 4
                    &\text{definition of $\hat q$}
                \end{array}
                $$

            \item $v_1 < \intdiv{B}{2}$:

                $$
                \begin{array}{llll}
                    2 (v_1 - 1) &\le&  B - 4
                    &\text{subclaim 5}

                    \\ \\

                    v_1 &\le& \frac{B}{2} - 1

                    \\ \\

                    v_1 &<& \intdiv{B}{2}
                    &\text{theorem~\ref{intdiv-theorem2}}
                \end{array}
                $$

        \end{enumerate}


        This concludes the proof of the contrapositive. Therefore the stated
        theorem is valid.
    \end{proof}
\end{theorem}

Remark: The proof of this theorem is correct but somehow unsatisfactory. It does
not give any insight because it is a collection of complex transformations. The
principal cause of complexity are nonlinear functions. The floor function
$\floor{x}$ is inherently nonlinear and the definition of $\hat q$ is nonlinear
(because it uses the nonlinear function \emph{min}). It might be possible to
gain more insight to concentrate on the difference between $\intdiv{u}{v}$ and
$\intdiv{u}{v_1 B^{n-1}}$ which is in the denominator. And then take advantage
of the fact $v_1 B^{n-1} \le v < v_1 B^n$.
